A Handbook for Tenured Track Faculty in Computer Science in the US

### 2024-07-13 demystify.tex
updates

\section{How long does to complete the CS PhD program?}\label{sec:time}

@@

Furthermore, avoid using Mr. or Mrs., or the professor's first name if you're not acquainted with them yet.  As you become more familiar with your prof and depending on their preferences, you may transition to addressing them by their first name.
For example, I prefer that my students and colleagues call me Vu. Some students call me \emph{Dr. Vu}, which I find a bit amusing but am totally fine with it.

@@

Note that in some universities the formal title Dr. Lastname is preferred over Prof. You just need to observe and follow the conventions at your particular institution. Additionally, be aware that not all faculty members may hold a PhD, in which case using Prof. Lastname is a suitable alternative.

@@

Students understandbly get excited about their upcoming PhD journey and want to buy new computer equipment and electronics to prepare. However, you should check with your prof. first.  They might have funding to buy you computers and other equipment (e.g., software, books, keyboards, headphones, tablets, etc). Many CS programs also provide budget or computer equipment to incoming PhD students (e.g., a laptop).

\section{Can I negotiate my PhD offer (e.g., if I have multiple offers)?}\label{sec:negotiate}

@@

\begin{table}
  \centering
  \small
  \caption{Top 50 CS PhD programs in the U.S. (\href{https://csrankings.org}{CSRankings}, Jan. 2024). \red{$^*$} indicates that the university has \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Viet-CS-Profs-US}{Vietnamese prof.} that can advise CS PhD students.}\label{tab:ranking}
  \begin{tabular}{rl|rl}
    \toprule
    1 & Carnegie Mellon & 26 & Duke University \\
    2 & Univ. of Illinois at Urbana-Champaign\red{$^*$}  & 27 & Univ. of California - Santa Barbara \\
    3 & Univ. of California-San Diego & 28 & Rutgers University\red{$^*$} \\
    4 & Georgia Institute of Technology & 29 & Univ. of California - Riverside\\
    5 & MIT                            & 30 & Pennsylvania State University  \\
    6 & Univ. of California - Berkeley& 31 & Northwestern University\\
    7 & University of Michigan - Ann Arbor\red{$^*$}   & 32& George Mason University\red{$^*$}\\
    8 & University of Washington      &33 &  Texas A\&M University\red{$^*$} \\
    9 &  Stanford University  &34&  University of Utah \\
    10 & Cornell University  & 35 &  North Carolina State University\\\
    11 & University of Maryland - College Park &  36& Ohio State \\
    12 & Northeastern University\red{$^*$} &37& University of Virginia  \\
    13 & Purdue University &38& Yale University \\
    14 & University of Wisconsin - Madison\red{$^*$} &39 & Univ. of California - Santa Cruz \\
    15 & University of Texas at Austin &40& Brown University \\
    16 & University of Pennsylvania &41 & Harvard University \\
    17 & Columbia University\red{$^*$} &42 & Boston University  \\
    18 & Princeton University\red{$^*$}  & 43& Rice University\\
    19 & New York University  & 44&  University at Buffalo\red{$^*$}\\
    20 &  University of Massachusetts-Amherst\red{$^*$} &45& University of Colorado-Boulder \\
    21 & Univ. of California - Los Angeles &46& University of Illinois at Chicago  \\
    22 & University of Southern California &47& University of Minnesota \\
    23 & University of Chicago &48& University of North Carolina\red{$^*$} \\
    24 & Stony Brook University\red{$^*$} &49& Arizona State University\red{$^*$} \\
    25 &  Univ. of California - Irvine&50& Univ. Of California - Davis \\
    \bottomrule
  \end{tabular}
\end{table}
% Virginia Tech\red{$^*$}  \\
\autoref{tab:ranking} lists the top 50 CS programs in the US from \href{https://www.csrankings.org}{CSRankings.org}, a ranking system  based on faculty publications at top CS conferences.

@@

\paragraph{History} This document was conceived during a lunch with a faculty at Mason.  We talked about why Mason was not able to attract good Vietnamese and other international students, despite having a much stronger CS program than many schools that these students want to go to (part of the reason is described in \autoref{sec:selecting-ranking-schools}). We wish there were a way for international students to know about the US PhD programs (also for US faculty to understand more about international students and therefore have a better chance of recruiting and working with them).

I was also a member of the large VietPhD group on Facebook and often browse Internet forums (e.g., \texttt{Reddit/gradadmission} and \texttt{GradCafe}). There I saw many questions from students about PhD programs.  However, most active participants are students in non-CS fields or not in the US, and their answers are unfortunately not always accurate and sometimes lead to more confusion. So I thought it would be useful to have a document that is specific to CS PhD programs in the US from an insider perspective.

@@

\paragraph{About Me} I am an associate professor in the CS dept at George Mason University. Before Mason, I was at the University of Nebraska-Lincoln (UNL). I have been in the PhD admission process at Mason and UNL for many years.  Currently I serve as the program director of the MS program in Software Engineering at Mason (thus also have experience with the MS admission process---which is quite different than PhD). My personal and lab website is at \href{https://dynaroars.github.io}{dynaroars.github.io}.

>>>>

\section{How long to complete the CS PhD program?}\label{sec:time}

@@

Furthermore, avoid using Mr., Mrs., Miss, or the professor's first name if you're not acquainted with them yet.  As you become more familiar with your prof and depending on their preferences, you may transition to addressing them by their first name.
For example, I prefer that my students and colleagues call me Vu. Some students call me \emph{Dr. Vu}, which I find a bit amusing but am totally fine with it.

@@

Note that in some universities the formal title Dr. Lastname is preferred over Prof. Moreover, be aware that not all faculty members may hold a PhD, in which case using Prof. Lastname is a suitable alternative. 	You just need to observe and follow the conventions at your particular institution.One way to determine how to address a professor is to observe how they sign their emails or how they introduce themselves in class. For example, I introduce myself as ``Vu'' in class.

@@

Students understandably get excited about their upcoming PhD journey and want to buy new computer equipment and electronics to prepare. However, you should check with your prof. first.  They might have funding to buy you computers and other equipment (e.g., software, books, keyboards, headphones, tablets, etc). Many CS programs also provide budget or computer equipment to incoming PhD students (e.g., a laptop).

\section{Negotiating PhD offer (e.g., having multiple offers)?}\label{sec:negotiate}

@@

\begin{table}[h]
  \centering
  \small
  \caption{The top 50 CS programs in the US from \href{https://www.csrankings.org}{CSRankings.org}, a ranking system  based on faculty publications at top CS conferences. (\href{https://csrankings.org}{CSRankings}, Jan. 2024). \red{$^*$} indicates that the university has \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Viet-CS-Profs-US}{Vietnamese prof.} that can advise CS PhD students.}\label{tab:ranking}
  \begin{tabular}{rl|rl}
    \toprule
    1 & Carnegie Mellon & 26 & Duke University \\
    2 & Univ. of Illinois at Urbana-Champaign\red{$^*$}  & 27 & Univ. of California - Santa Barbara \\
    3 & Univ. of California-San Diego & 28 & Rutgers University\red{$^*$} \\
    4 & Georgia Institute of Technology & 29 & Univ. of California - Riverside\\
    5 & MIT                            & 30 & Pennsylvania State University  \\
    6 & Univ. of California - Berkeley& 31 & Northwestern University\\
    7 & University of Michigan - Ann Arbor\red{$^*$}   & 32& George Mason University\red{$^*$}\\
    8 & University of Washington      &33 &  Texas A\&M University\red{$^*$} \\
    9 &  Stanford University  &34&  University of Utah \\
    10 & Cornell University  & 35 &  North Carolina State University\\\
    11 & University of Maryland - College Park &  36& Ohio State \\
    12 & Northeastern University\red{$^*$} &37& University of Virginia  \\
    13 & Purdue University &38& Yale University \\
    14 & University of Wisconsin - Madison\red{$^*$} &39 & Univ. of California - Santa Cruz \\
    15 & University of Texas at Austin &40& Brown University \\
    16 & University of Pennsylvania &41 & Harvard University \\
    17 & Columbia University\red{$^*$} &42 & Boston University  \\
    18 & Princeton University\red{$^*$}  & 43& Rice University\\
    19 & New York University  & 44&  University at Buffalo\red{$^*$}\\
    20 &  University of Massachusetts-Amherst\red{$^*$} &45& University of Colorado-Boulder \\
    21 & Univ. of California - Los Angeles &46& University of Illinois at Chicago  \\
    22 & University of Southern California &47& University of Minnesota \\
    23 & University of Chicago &48& University of North Carolina\red{$^*$} \\
    24 & Stony Brook University\red{$^*$} &49& Arizona State University\red{$^*$} \\
    25 &  Univ. of California - Irvine&50& Univ. Of California - Davis \\
    \bottomrule
  \end{tabular}
\end{table}
% Virginia Tech\red{$^*$}  \\

%\autoref{tab:ranking} lists .

@@

\paragraph{History} This document was conceived during a lunch with a faculty at Mason.  We talked about why Mason was not able to attract good Vietnamese and other international students, despite having a much stronger CS program than many schools that these students want to go to (part of the reason is described in \autoref{sec:selecting-ranking-schools}). We wish there were a way for international students to know about PhD programs in the US.

I was also a member of the large VietPhD group on Facebook and often browse Internet forums (e.g., \texttt{Reddit/gradadmission} and \texttt{GradCafe}). There I saw many questions from students about PhD programs.  However, most participants are students in non-CS fields or not in the US, and their answers are not always accurate and can lead to more confusion. So I thought it would be useful to have a document that is specific to CS PhD programs in the US from an insider perspective.

@@

\paragraph{About Me} I am an associate professor in the CS dept at George Mason University (Mason). Before Mason, I was at the University of Nebraska-Lincoln (UNL). I have been in the PhD admission process at Mason and UNL for many years.  I also serve as the program director of the MS program in Software Engineering at Mason (thus also have experience with the MS admission process---which is quite different than PhD). My personal and lab website is at \href{https://dynaroars.github.io}{dynaroars.github.io}.

### 2024-07-13 faculty.tex
updates

@@

\section{Types of Funding}
\paragraph{NSF} NSF is the obvious funding source. Among many good things about NSF, two stand out: (transparency) proposals are evaluated by an independent panel and are based on merit rather than personal connections; (autonomy) the flexibility and freedom given to focus on pure research without the pressure of producing concrete deliverables.  Because of these reasons, NSF funding is often considered the prestigious and competitive, even for very experienced senior researchers from top places (which don't mean anything). Note that a PI can use at most 2 months of summer salary from NSF grants (if you need the 3rd summer month, you need to find other sources of funding).

\paragraph{DARPA, ONR, and other defense sponsors} DARPA, ONR, Army, Airforce, etc are like a contract, with deliverables required at regular intervals. Contracts are awarded in phases to multiple teams, and funding is cut for underperforming teams. This puts a lot of pressure on both the PIs and their students, requiring them to spend significant time on reports, meetings, and product development (e.g., creating hacker-proof software rather than research prototypes). Additionally, in many cases funding often depends on existing connections or being affiliated with institutions that frequently receive such awards.

\subsection{Internal Funding}

\subsection{Industrial Gifts}
As its name suggests, industrial gifts are unrestricted research funding from companies such as Amazon, Microsoft, Facebook, and Google. These are highly competitive and prestigious (e.g., the solicitation is open worldwide, and you compete with top researchers from top places).  They are the best kind of funding for three reasons: (1) indirect costs or overheads are typically not levied on these gifts (so you get most of the money), (2) there is no associated timeline by when you should spend them (they give it to you and forget about it), and (3) you can use them for a broader range of things than university and grant/contract funding (e.g., fun events for your research group).

\section{My Story}

@@

Successfully mentoring students is a key and most rewarding part of your job as a faculty member. 
There will be times when you feel discouraged and frustrated, but there will also be times when you feel proud and happy.  
You will be expected to mentor undergraduate students and graduate students. You will likely need to graduate at least one PhD student to get tenure.

@@

\paragraph{PhD Dissertation Committee} Being in a PhD committee is a common task to help your colleagues (e.g., the adviser of the student asks you to be in student dissertation committee) and the students (e.g., you are involved or are interested in the research of the student).  You can serve in the committee of students in your dept, in other dept, or even in a different university (external committee member). This is a good way to build your network (and also help your students build their network).

@@

I did not do these visits, partly because I am not a travel person, and also I felt that I would not have problems finding letter writers.
Instead, during the Spring'23 semester before my tenure application, I put lots of effort in helping the university in recruiting faculty (by meeting with 20+ faculty candidates), even though I am not in the search committee and really didn't have service requirements during this pre-tenure study break. This helps me build a good relationship with my colleagues in the dept.

\section{3rd year Renewal Package}

@@

Toward the end of your 3rd year (or early 4th year), you will need to submit a ``midterm'' \emph{renewal} application to renew your contract for the 4th year (and perhaps also 5th year).
While somewhat similar to tenure evaluation, this is less intense and  does not require reference letters or presentation.
%The department will then review your application and vote on whether to renew you for your 4th (and perhaps also 5th year). 
Your mentor will also present and argue for your case in the dept. meeting, but the decision is made within the dept. and usually does not go up to the college or university.  It is still signed by the Dean, who often just follows the dept's recommendation.

While the renewal is straightforward for most people, it can be surprise and shock to some. Ideally, it should not be a surprise, as your annual evaluation should give you some idea on how you are doing, e.g., if you are not doing well, the annual evaluation should tell you that you need to improve. However, it can be a surprise due to various reasons such as new dept. chair or administrative changes who have different expectations.

>>>>

\section{Types of Funding}
\paragraph{NSF} NSF is the obvious funding source for CS. 
Two things stand out for NSF grants: (transparency) proposals are evaluated by an independent panel and are based on merit rather than personal connections; (autonomy) the flexibility and freedom given to focus on pure research without the pressure of producing concrete deliverables.  Because of these reasons, NSF funding is often considered the prestigious and competitive, even for very experienced senior researchers from top places (which don't mean anything). Note that a PI can use at most 2 months of summer salary from NSF grants (if you need the 3rd summer month, you need to find other sources of funding).

\paragraph{DARPA, ONR, and other defense sponsors} Most of these awards are like a \emph{contract}, with deliverables required at regular intervals. Contracts are awarded in phases to multiple teams, and funding is cut for underperforming teams. This puts a lot of pressure on both the PIs and their students, requiring them to spend significant time on reports, meetings, and product development (e.g., creating hacker-proof software rather than research prototypes). Additionally, in many cases funding often depends on existing connections or being affiliated with institutions that frequently receive such awards.

\subsection{Internal Funding}
These are funds from your university, city, or state (e.g., Virginia has the CCI program that provide ``seed'' grants for researchers from universities in Virginia).  Also, your ``startup'' package is also an internal funding to kickstart your research. These funding are less competitive than NSF, but they are also less prestigious (e.g., they aren't much and usually don't count as much for tenure).  However, they are good sources of funding to get preliminary results and later apply for traditional grants (and many internal funding programs explicitly encourage this).

\subsection{Industrial Gifts}
Industrial gifts are \emph{unrestricted funding} from companies such as Amazon, Microsoft, Facebook, and Google. These are highly competitive and prestigious (e.g., the solicitation is open worldwide, and you compete with top researchers from top places).  They are the best kind of funding for three reasons: (1) indirect costs or overheads are typically not levied on these gifts (so you get most of the money), (2) there is no associated timeline by when you should spend them (they give it to you and forget about it), and (3) you can use them for a broader range of things than university and grant/contract funding (e.g., summary salary, equipments, or even bowling events for your group).

\paragraph{Personal Experiences}

@@

@@

Successfully mentoring students is a key and most rewarding part of your job as a faculty member. 
There will be times when you feel discouraged and frustrated, but there will also be times when you feel proud and happy.  
For tenure you will be expected to mentor undergraduate students and graduate students. You will likely need to graduate at least one PhD student to get tenure.

@@

\paragraph{PhD Dissertation Committee} Being in a PhD committee is a common task to help your colleagues (e.g., the adviser of the student asks you to be in student dissertation committee) and the students (e.g., you are involved or are interested in the research of the student).  You can serve in the PhD or MS committee of students in your dept, in other dept, or even in a different university (external committee member).

@@

I did not do these visits, partly because I am not a travel person, and also I felt that I would have enough good letter writers.
However, during the Spring'23 semester before my tenure application, I did participate more in the dept. activities, e.g., in recruiting by meeting with 20+ faculty candidates, even though I am not in the search committee and really didn't have service requirements during this pre-tenure study break. This helps me build a good relationship with my colleagues in the dept.

\section{Midterm Review}

@@

Midterm review happens midway through the tenure period, i.e., the 3rd year in a 6-year tenure track. You will need to submit a midterm \emph{renewal} application to renew your contract (and your next renewal will be the tenure application).
While somewhat similar to tenure evaluation described above, midterm review is less intense and  does not require reference letters or presentation.
%The department will then review your application and vote on whether to renew you for your 4th (and perhaps also 5th year). 
Your mentor will also present and argue for your case in the dept. meeting, but the decision is made within the dept. and usually does not go up to the college or university.  It is still signed by the Dean, who often just follows the dept's recommendation.

While the renewal is straightforward for most people, it can be a surprise and shock to some. Ideally, it should not be a surprise, as your annual evaluation or dept. chair and mentor should tell you if you're not doing well and what to improve.  However, things can become complicated due to various reasons such as new dept. chair or administrative changes who have different expectations.
